---
title: "Plasmid Triage Analysis"
subtitle: "MOB-suite + AMRFinderPlus Combined Output summary and visualization"
author: "Shawn Hawken"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This notebook combines MOB-recon contig classification with AMRFinderPlus resistance
gene calls to:
  1. triage short-read assemblies for long-read sequencing based on:

- Presence of potential plasmids carrying AMR genes
- Presence of AMR genes on chromosome-designated contigs that may be misclassified plasmids
  2. Generate or support hypotheses regarding presence of plasmid and AMR sequences in short-tread data  

Note & Disclaimer: this workflow is a screening workflow. We prioritize sensitivity over specificity — ambiguous 
calls are retained as potential plasmid detections to avoid false negatives. Plasmid presence/absence inferences should always be interpreted within the clinical and epidemiological context of the isolates in the investigation. For surveillance purposes only, not for diagnostic or treatment.

---

## Load Libraries
```{r libraries}
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

---

## Load Data

> **Edit the file paths below** to point to your combined output files.

```{r load-data} 

contigs <- read.delim(
  "../your_combined_contig_report.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

amr <- read.delim(
  "your_combined_amrfinder.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

```

---

## Data Summary
```{r summary}

cat("Total contigs:       ", nrow(contigs), "\n")
cat("Plasmid contigs:     ", sum(contigs$molecule_type == "plasmid"), "\n")
cat("Chromosome contigs:  ", sum(contigs$molecule_type == "chromosome"), "\n")
cat("Samples:             ", length(unique(contigs$sample)), "\n")
cat("Total AMR hits:      ", nrow(amr), "\n")
cat("Unique AMR genes across samples:       ", length(unique(amr$Element.name)), "\n")

```

--- 
## Summary of AMR, Virulence and Stress genes across choromosome and plasmid designated sequence

```{r}
kable(table(amr$Element.symbol,amr$Type))
```


---

## Table 1 — Plasmid Bins with AMR Genes

One row per plasmid bin per sample, sorted by total bp descending within each sample.
Only bins classified as `plasmid` by MOB-recon with at least one AMR hit are shown.
```{r plasmid-amr-table}

# Summarize plasmid contigs to bin level
plasmid_contigs <- contigs[contigs$molecule_type == "plasmid", ]

bin_size <- plasmid_contigs %>%
  group_by(sample, primary_cluster_id) %>%
  summarise(
    total_bp   = sum(size, na.rm = TRUE),
    total_kb   = round(sum(size, na.rm = TRUE) / 1000, 2),
    n_contigs  = n(),
    rep_type   = paste(unique(rep_type.s.[rep_type.s. != "-"]), collapse = "; "),
    mobility   = paste(unique(predicted_mobility[predicted_mobility != "-"]), collapse = "; "),
    .groups    = "drop"
  )

# Summarize AMR hits for plasmid bins
# Strip plasmid_ prefix from contig_bin to match primary_cluster_id
plasmid_amr <- amr[amr$molecule_type == "plasmid", ] %>%
  filter(!is.na(Element.symbol)) %>%
  mutate(primary_cluster_id = sub("^plasmid_", "", contig_bin)) %>%
  group_by(sample, primary_cluster_id) %>%
  summarise(
    n_amr_hits  = n(),
    amr_genes   = paste(sort(unique(Element.symbol)), collapse = "; "),
    amr_classes = paste(sort(unique(Class[!is.na(Class)])), collapse = "; "),
    .groups     = "drop"
  )

# Join on both sample and primary_cluster_id — only bins with AMR hits
plasmid_bin_amr <- bin_size %>%
  inner_join(plasmid_amr, by = c("sample", "primary_cluster_id")) %>%
  arrange(sample, desc(total_bp))

kable(
  plasmid_bin_amr[, c("sample", "primary_cluster_id", "rep_type", "mobility",
                       "n_contigs", "total_kb", "n_amr_hits", "amr_genes", "amr_classes")],
  caption   = "Plasmid bins with AMR gene detections — sorted by size within sample",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "800px")
```




---

## Table 2 — Chromosome Contigs with AMR Genes

One row per sample, one column per unique AMR gene (hit count).
Chromosome-designated contigs may represent misclassified plasmids in short-read assemblies.
```{r chromosome-amr-table}

chrom_amr <- amr[amr$molecule_type == "chromosome", ] %>%
  filter(!is.na(Element.symbol))

# Wide format: one row per sample, one column per unique AMR gene (counts)
chrom_amr_wide <- chrom_amr %>%
  group_by(sample, Element.symbol) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from  = Element.symbol,
    values_from = count,
    values_fill = 0
  ) %>%
  arrange(sample)

# Add row sums and column sums
gene_cols <- setdiff(colnames(chrom_amr_wide), "sample")

chrom_amr_wide_display <- chrom_amr_wide %>%
  mutate(row_total = rowSums(across(all_of(gene_cols))))

col_totals <- chrom_amr_wide_display %>%
  summarise(
    sample    = "column_total",
    across(all_of(gene_cols), sum),
    row_total = sum(row_total)
  )

chrom_amr_wide_display <- bind_rows(chrom_amr_wide_display, col_totals)

kable(
  chrom_amr_wide_display,
  caption   = "AMR gene counts on chromosome-designated contigs per sample (with row and column totals)",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(nrow(chrom_amr_wide_display), bold = TRUE, background = "#f0f0f0") %>%
  scroll_box(height = "400px")

```



## Summary of plasmid AMR hits on both plasmid designated contigs OR chromosme designated contigs

```{r}

cat("Samples with plasmid AMR (linked on the same contigs):    ", length(unique(plasmid_bin_amr$sample)), "\n")
cat("Samples with Chromosomal AMR (Examine samples for genes that could be on plasmids):      ", length(unique(chrom_amr$sample)), "\n")

```
---

## Literature-Based Flagging

 **Placeholder COMING SOON** — add gene symbols or classes of interest here for manual review 
 prioritization. Examples: carbapenemases (blaNDM, blaKPC, blaOXA-48), ESBLs (blaCTX-M), 
 colistin (mcr).
 
```{r}


# Add gene symbols to flag here
flag_genes <- c(
  # "blaNDM", "blaKPC", "blaOXA-48", "blaCTX-M", "mcr"
)

if (length(flag_genes) > 0) {
  plasmid_bin_amr$priority_flag <- sapply(plasmid_bin_amr$amr_genes, function(g) {
    any(sapply(flag_genes, function(f) grepl(f, g, ignore.case = TRUE)))
  })
  chrom_amr_wide$priority_flag <- sapply(chrom_amr_wide$sample, function(s) {
    hits <- chrom_amr$Element.symbol[chrom_amr$sample == s]
    any(sapply(flag_genes, function(f) any(grepl(f, hits, ignore.case = TRUE))))
  })
} else {
  message("No literature flags defined — skipping flagging step.")
}

```

---

## Auspice Metadata Output

Wide-format dataframe for Auspice visualization. One row per sample, one column per
unique plasmid or chromosome AMR fingerprint. Column names are constructed as:
`{molecule_type}_{primary_cluster_id}_{rep_type}_{amr_genes_sorted}`.
Cell values are total coverage (in bp) of that bin in that sample.
AMR genes are sorted alphabetically within each fingerprint so the same plasmid
produces the same column name across different samples.  
*Note:* you will need to merge this so the first column of the dataframe matches up with ID names on your phylogenetic tree before viewing with auspice.  

```{r}

# --- Plasmid columns ---
# Fingerprint: plasmid_{primary_cluster_id}_{rep_type}_{sorted_amr_genes}
plasmid_auspice <- plasmid_bin_amr %>%
  mutate(
    rep_type_clean = ifelse(rep_type == "" | is.na(rep_type), "-", rep_type),
    fingerprint    = paste("plasmid", primary_cluster_id, rep_type_clean, amr_genes, sep = "_")
  ) %>%
  select(sample, fingerprint, total_bp) %>%
  pivot_wider(
    names_from  = fingerprint,
    values_from = total_bp,
    values_fill = 0
  )

# --- Chromosome columns ---
# Fingerprint: chromosome_chromosome_-_{gene}
chrom_auspice <- chrom_amr %>%
  group_by(sample, Element.symbol) %>%
  summarise(
    # Use total bp of contigs carrying this gene as the value
    hit_count = n(),
    .groups   = "drop"
  ) %>%
  mutate(fingerprint = paste("chromosome", "chromosome", "-", Element.symbol, sep = "_")) %>%
  select(sample, fingerprint, hit_count) %>%
  pivot_wider(
    names_from  = fingerprint,
    values_from = hit_count,
    values_fill = 0
  )

# --- Combine into one wide dataframe ---
all_samples <- data.frame(sample_id = unique(contigs$sample))

auspice_df <- all_samples %>%
  left_join(plasmid_auspice, by = c("sample_id" = "sample")) %>%
  left_join(chrom_auspice,   by = c("sample_id" = "sample"))

# Replace NAs with 0 for samples with no hits
auspice_df[is.na(auspice_df)] <- 0

```



```{r}
# Write out
write.csv(auspice_df, "plasmid_triage_auspice_metadata.csv", row.names = FALSE)

cat("Auspice metadata written to plasmid_triage_auspice_metadata.csv\n")
cat("Rows (samples):         ", nrow(auspice_df), "\n")
cat("Total columns:          ", ncol(auspice_df), "\n")
cat("Plasmid columns:        ", ncol(plasmid_auspice) - 1, "\n")
cat("Chromosome AMR columns: ", ncol(chrom_auspice) - 1, "\n")
```


## Summarize samples with potential plasmids and AMR genes in common
-summarize amr detection and potential plasmid detection across samples included in the input data  
-This helps evaluate if there are samples that potentially share plasmid content (ie supports hypothesis that this could be due to plasmid transfer or transmission)   


### Plasmid Fingerprint Summary
```{r plasmid-fingerprint-summary}

plasmid_fingerprints

# Build full fingerprint: primary_cluster_id + sorted amr_genes
plasmid_fingerprints <- plasmid_bin_amr %>%
  mutate(
    rep_type_clean = ifelse(rep_type == "" | is.na(rep_type), "-", rep_type),
    fingerprint    = paste(primary_cluster_id, amr_genes, sep = "_")
  )


plasmid_fingerprints$ID<-plasmid_fingerprints$sample


# Summary table: one row per fingerprint, count and list of samples
plasmid_fingerprint_summary <- plasmid_fingerprints %>%
  group_by(fingerprint, primary_cluster_id, rep_type_clean, amr_genes, amr_classes) %>%
  summarise(
    n_samples   = n_distinct(ID),
    samples     = paste(sort(unique(ID)), collapse = "; "),
    .groups     = "drop"
  ) %>%
  arrange(desc(n_samples))

kable(
  plasmid_fingerprint_summary[, c("fingerprint", "primary_cluster_id", "rep_type_clean",
                                   "amr_genes", "amr_classes", "n_samples", "samples")],
  caption   = "Plasmid fingerprints detected across samples",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")


write.csv(plasmid_fingerprint_summary, "2026-05-21_plasmid_fingerprint_summary.csv", row.names = FALSE)

```



### Plasmid Fingerprint Presence/Absence Heatmap
```{r plasmid-heatmap}
# Build presence/absence matrix
plasmid_matrix <- plasmid_fingerprints %>%
  mutate(present = 1) %>%
  select(ID, fingerprint, present) %>%
  pivot_wider(
    names_from  = fingerprint,
    values_from = present,
    values_fill = 0
  )

# Save as dataframe for exploration
plasmid_matrix_df <- as.data.frame(plasmid_matrix)

# Helper to cluster and plot plasmid heatmap — fingerprints on y, samples on x
plot_plasmid_heatmap <- function(mat_df, title) {
  mat_numeric <- as.matrix(mat_df[, -1])
  rownames(mat_numeric) <- mat_df$ID

  # Guard: need at least 2 rows and 2 columns to cluster
  if (nrow(mat_numeric) < 2 | ncol(mat_numeric) < 2) {
    message("Not enough samples or fingerprints to cluster — skipping: ", title)
    return(NULL)
  }

  sample_order <- rownames(mat_numeric)[hclust(dist(mat_numeric))$order]
  fp_order     <- colnames(mat_numeric)[hclust(dist(t(mat_numeric)))$order]

  mat_df %>%
    pivot_longer(cols = -ID, names_to = "fingerprint", values_to = "present") %>%
    mutate(
      sample      = factor(ID,      levels = sample_order),
      fingerprint = factor(fingerprint, levels = fp_order)
    ) %>%
    ggplot(aes(x = ID, y = fingerprint, fill = factor(present))) +
    geom_tile(color = "white") +
    scale_fill_manual(
      values = c("0" = "#f0f0f0", "1" = "#2c7bb6"),
      labels = c("0" = "absent",  "1" = "present"),
      name   = ""
    ) +
    theme_minimal() +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y     = element_text(size = 8),
      panel.grid      = element_blank(),
      legend.position = "top"
    ) +
    labs(title = title, x = "Sample", y = "Plasmid fingerprint (bin + AMR genes)")
}

# Version 1 — all fingerprints
plot_plasmid_heatmap(
  plasmid_matrix_df,
  "Plasmid fingerprint presence/absence — all samples (clustered)"
)

# Version 2 — drop fingerprints present in only 1 sample
multi_sample_fps <- names(which(colSums(plasmid_matrix_df[, -1]) > 1))

if (length(multi_sample_fps) == 0) {
  message("No plasmid fingerprints shared across more than one sample.")
} else {
  plasmid_matrix_df_shared <- plasmid_matrix_df[, c("ID", multi_sample_fps)]
  plot_plasmid_heatmap(
    plasmid_matrix_df_shared,
    "Plasmid fingerprint presence/absence — shared across >1 sample (clustered)"
  )
}


write.csv(plasmid_matrix_df, "plasmid_matrix_summary.csv", row.names = FALSE)

```

## Chromosome AMR genes presence/absence heatmaps
```{r chromosome-heatmap}
# Build count matrix
chrom_matrix <- chrom_amr %>%
  filter(!is.na(Element.symbol)) %>%
  group_by(sample, Element.symbol) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from  = Element.symbol,
    values_from = count,
    values_fill = 0
  )

# Save as dataframe for exploration
chrom_matrix_df <- as.data.frame(chrom_matrix)

# Helper to cluster and plot chromosome heatmap
plot_chrom_heatmap <- function(mat_df, title) {
  mat_numeric <- as.matrix(mat_df[, -1])
  rownames(mat_numeric) <- mat_df$sample

  # Guard: need at least 2 rows and 2 columns to cluster
  if (nrow(mat_numeric) < 2 | ncol(mat_numeric) < 2) {
    message("Not enough samples or genes to cluster — skipping: ", title)
    return(NULL)
  }

  sample_order <- rownames(mat_numeric)[hclust(dist(mat_numeric))$order]
  gene_order   <- colnames(mat_numeric)[hclust(dist(t(mat_numeric)))$order]

  mat_df %>%
    pivot_longer(cols = -sample, names_to = "gene", values_to = "count") %>%
    mutate(
      sample = factor(sample, levels = sample_order),
      gene   = factor(gene,   levels = gene_order)
    ) %>%
    ggplot(aes(x = gene, y = sample, fill = count)) +
    geom_tile(color = "white") +
    scale_fill_gradient(
      low  = "#f0f0f0",
      high = "#d7191c",
      name = "hit count"
    ) +
    theme_minimal() +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1, size = 8),
      axis.text.y     = element_text(size = 8),
      panel.grid      = element_blank(),
      legend.position = "top"
    ) +
    labs(title = title, x = "AMR gene", y = "Sample")
}

# Version 1 — all genes
plot_chrom_heatmap(
  chrom_matrix_df,
  "Chromosome AMR gene counts — all samples (clustered)"
)

# Version 2 — drop genes present in only 1 sample
multi_sample_genes <- names(which(colSums(chrom_matrix_df[, -1] > 0) > 1))

if (length(multi_sample_genes) == 0) {
  message("No chromosome AMR genes shared across more than one sample.")
} else {
  chrom_matrix_df_shared <- chrom_matrix_df[, c("sample", multi_sample_genes)]
  plot_chrom_heatmap(
    chrom_matrix_df_shared,
    "Chromosome AMR gene counts — shared across >1 sample (clustered)"
  )
}

```


## Plot heatmaps by phylogenetic trees if tree present 
-This is used to evaluate tree side by side with the matrix plot to examine phylogenetic distribution of detected genes and plasmids  

```{r}
# read in tree
gubbins.tree<-read.tree("tree.nwk")
#clean up tree tip names if needed to match the matrix you wish to plot
#gubbins.tree$tip.label<-sub("_T1","",gubbins.tree$tip.label) #cleanup tip labels to match matrix rownames


# trim tree to only include tips that are in the distance matrix you wish to analyze
tips<-setdiff(gubbins.tree$tip.label,rownames(chrom_matrix_df))

trimmed_tree <- drop.tip(gubbins.tree, tips, trim.internal = TRUE)

plotTree(trimmed_tree) # this is the tree used to order the heatmap


# plot the resulting presence-absence matrix heatmap
# note the tip.labels and row names on the matrix must be the same!
#order matrix by the tree tips

ordered_mat<-chrom_matrix_df[rev(trimmed_tree$tip.label),]

# remove genes that are not present in any of the tree members
ordered_mat<-ordered_mat[,colSums(ordered_mat)>0]

#plot ordered matrix
pheatmap(ordered_mat,
         legend_breaks = c(0,1), 
         legend_labels = c("Absent","Present"),
         annotation_names_row = TRUE,
         annotation_names_col = TRUE,
         cluster_rows = FALSE,
         cluster_cols=TRUE,
         color = (c("lightblue","red")),
          )

#evaluate tree side by side with the matrix plot to examine phylogenetic distribution of detected genes and plasmids
```

