---
title: "Plasmid Triage Analysis"
subtitle: "MOB-suite + AMRFinderPlus Combined Output"
author: "Shawn Hawken"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This notebook combines MOB-recon contig classification with AMRFinderPlus resistance
gene calls, linking them by contig ID to answer questions like:

- Which Inc groups are carrying which resistance genes?
- Are carbapenemase genes on conjugative plasmids?
- Which plasmid bins carry multiple resistance genes?

---

## Load Libraries

```{r libraries}
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

## Load Data

> **Edit the file paths below** to point to your combined output files.

```{r load-data}
contigs <- read.delim(
  "combined_contig_report.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

amr <- read.delim(
  "combined_amrfinder.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)
```

**Note:** R replaces spaces and special characters in column names with `.` when
reading tab-delimited files — e.g. `Contig id` becomes `Contig.id` and
`Gene symbol` becomes `Gene.symbol`.

---

## Filter to Plasmid Contigs

```{r filter-plasmids}
plasmids <- contigs[contigs$molecule_type == "plasmid", ]

cat("Total contigs:", nrow(contigs), "\n")
cat("Plasmid contigs:", nrow(plasmids), "\n")
cat("Samples:", length(unique(plasmids$sample)), "\n")
```

---

## Join AMR Hits to Plasmid Metadata

```{r join-data}
plasmid_amr <- merge(
  plasmids, amr,
  by.x  = c("sample", "contig_id"),
  by.y  = c("sample", "Contig.id"),
  all.x = TRUE
)
```

---

## Summary Tables

### Plasmid Bins per Sample

```{r bins-per-sample}
bins_per_sample <- aggregate(
  cbind(plasmid_bins = primary_cluster_id) ~ sample,
  data = plasmids,
  FUN  = function(x) length(unique(x))
)
bins_per_sample$total_contigs <- ave(
  plasmids$sample, plasmids$sample, FUN = length
)[match(bins_per_sample$sample, plasmids$sample)]
bins_per_sample <- bins_per_sample[order(-bins_per_sample$plasmid_bins), ]

kable(bins_per_sample, caption = "Plasmid bins identified per sample", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Total Plasmid Size per Bin

```{r plasmid-size}
plasmid_sizes <- aggregate(
  contig_size ~ sample + primary_cluster_id,
  data = plasmids,
  FUN  = sum
)
names(plasmid_sizes)[names(plasmid_sizes) == "contig_size"] <- "total_size_bp"

plasmid_sizes$n_contigs <- ave(
  plasmids$contig_size, 
  paste(plasmids$sample, plasmids$primary_cluster_id),
  FUN = length
)[match(
  paste(plasmid_sizes$sample, plasmid_sizes$primary_cluster_id),
  paste(plasmids$sample, plasmids$primary_cluster_id)
)]

plasmid_sizes$rep_type <- tapply(plasmids$rep_type,
  paste(plasmids$sample, plasmids$primary_cluster_id),
  FUN = function(x) paste(unique(x[!is.na(x)]), collapse = "; ")
)[match(
  paste(plasmid_sizes$sample, plasmid_sizes$primary_cluster_id),
  names(tapply(plasmids$rep_type, paste(plasmids$sample, plasmids$primary_cluster_id), FUN = function(x) x[1]))
)]

plasmid_sizes$mobility <- tapply(plasmids$predicted_mobility,
  paste(plasmids$sample, plasmids$primary_cluster_id),
  FUN = function(x) paste(unique(x[!is.na(x)]), collapse = "; ")
)[match(
  paste(plasmid_sizes$sample, plasmid_sizes$primary_cluster_id),
  names(tapply(plasmids$predicted_mobility, paste(plasmids$sample, plasmids$primary_cluster_id), FUN = function(x) x[1]))
)]

plasmid_sizes <- plasmid_sizes[order(plasmid_sizes$sample, -plasmid_sizes$total_size_bp), ]

kable(plasmid_sizes, caption = "Estimated plasmid size per bin (sum of contig lengths)", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Inc Groups by Resistance Class

```{r inc-amr-table}
amr_with_genes <- plasmid_amr[!is.na(plasmid_amr$Gene.symbol), ]
inc_amr_long   <- as.data.frame(table(amr_with_genes$rep_type, amr_with_genes$Class))
names(inc_amr_long) <- c("rep_type", "Class", "n")
inc_amr <- reshape(inc_amr_long, idvar = "rep_type", timevar = "Class",
                   direction = "wide")
names(inc_amr) <- sub("^n\\.", "", names(inc_amr))
inc_amr[is.na(inc_amr)] <- 0

kable(inc_amr, caption = "Inc group vs. AMR class (count of AMR hits)", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## Visualizations

### AMR Genes per Resistance Class

```{r amr-class-plot, fig.width=8, fig.height=5}
amr_with_class <- plasmid_amr[!is.na(plasmid_amr$Class), ]
amr_counts     <- as.data.frame(table(Class = amr_with_class$Class))
amr_counts     <- amr_counts[order(-amr_counts$Freq), ]

ggplot(amr_counts, aes(x = reorder(Class, Freq), y = Freq, fill = Class)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "AMR Hits by Resistance Class",
    x     = "Resistance Class",
    y     = "Number of Hits"
  ) +
  theme_minimal()
```

### Mobility Class Distribution

```{r mobility-plot, fig.width=6, fig.height=4}
plasmids_mob    <- plasmids[!is.na(plasmids$predicted_mobility), ]
mobility_counts <- as.data.frame(table(predicted_mobility = plasmids_mob$predicted_mobility))

ggplot(mobility_counts, aes(x = predicted_mobility, y = Freq, fill = predicted_mobility)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Plasmid Mobility Class Distribution",
    x     = "Mobility",
    y     = "Number of Contigs"
  ) +
  theme_minimal()
```

### Plasmid Bin Sizes

```{r size-plot, fig.width=8, fig.height=5}
ggplot(plasmid_sizes, aes(x = reorder(primary_cluster_id, total_size_bp),
                          y = total_size_bp / 1000,
                          fill = sample)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Estimated Plasmid Size per Bin",
    x     = "Plasmid Bin",
    y     = "Estimated Size (kb)",
    fill  = "Sample"
  ) +
  theme_minimal()
```

---

## AMR on Conjugative Plasmids

Flagging AMR hits found on conjugative plasmid bins — highest transmission risk.

```{r conjugative-amr}
conjugative_amr <- plasmid_amr[
  !is.na(plasmid_amr$predicted_mobility) &
  plasmid_amr$predicted_mobility == "conjugative" &
  !is.na(plasmid_amr$Gene.symbol), 
  c("sample", "primary_cluster_id", "rep_type", "Gene.symbol", "Class", "Subclass",
    "X..Coverage.of.reference.sequence", "X..Identity.to.reference.sequence")
]
names(conjugative_amr)[names(conjugative_amr) == "X..Coverage.of.reference.sequence"] <- "coverage"
names(conjugative_amr)[names(conjugative_amr) == "X..Identity.to.reference.sequence"]  <- "identity"
conjugative_amr <- conjugative_amr[order(conjugative_amr$sample, conjugative_amr$primary_cluster_id), ]

if (nrow(conjugative_amr) > 0) {
  kable(conjugative_amr, caption = "AMR genes on conjugative plasmids", row.names = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
} else {
  cat("No AMR genes detected on conjugative plasmids.\n")
}
```

---

## Session Info

```{r session-info}
sessionInfo()
```
