---
title: "Plasmid Triage Analysis"
subtitle: "MOB-suite + AMRFinderPlus Combined Output"
author: "Shawn Hawken"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This notebook combines MOB-recon contig classification with AMRFinderPlus resistance
gene calls to triage short-read assemblies for long-read sequencing based on:

- Presence of potential plasmids carrying AMR genes
- Presence of AMR genes on chromosome-designated contigs that may be misclassified plasmids

Note: this workflow is a screen. We prioritize sensitivity over specificity — ambiguous 
calls are retained as potential plasmid detections to avoid false negatives.

---

## Load Libraries
```{r libraries}
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
```

---

## Load Data

> **Edit the file paths below** to point to your combined output files.

```{r load-data} 

contigs <- read.delim(
  "../Raabe_et_al_test/combined_contig_report.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

amr <- read.delim(
  "../Raabe_et_al_test/combined_amrfinder.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

```

---

## Data Summary
```{r summary}

cat("Total contigs:       ", nrow(contigs), "\n")
cat("Plasmid contigs:     ", sum(contigs$molecule_type == "plasmid"), "\n")
cat("Chromosome contigs:  ", sum(contigs$molecule_type == "chromosome"), "\n")
cat("Samples:             ", length(unique(contigs$sample)), "\n")
cat("Total AMR hits:      ", nrow(amr), "\n")

```

---

## Table 1 — Plasmid Bins with AMR Genes

One row per plasmid bin per sample, sorted by total bp descending within each sample.
Only bins classified as `plasmid` by MOB-recon with at least one AMR hit are shown.
```{r plasmid-amr-table}

# Summarize plasmid contigs to bin level
plasmid_contigs <- contigs[contigs$molecule_type == "plasmid", ]

bin_size <- plasmid_contigs %>%
  group_by(sample, primary_cluster_id) %>%
  summarise(
    total_bp   = sum(size, na.rm = TRUE),
    total_kb   = round(sum(size, na.rm = TRUE) / 1000, 2),
    n_contigs  = n(),
    rep_type   = paste(unique(rep_type.s.[rep_type.s. != "-"]), collapse = "; "),
    mobility   = paste(unique(predicted_mobility[predicted_mobility != "-"]), collapse = "; "),
    .groups    = "drop"
  )

# Summarize AMR hits for plasmid bins
# Strip plasmid_ prefix from contig_bin to match primary_cluster_id
plasmid_amr <- amr[amr$molecule_type == "plasmid", ] %>%
  filter(!is.na(Element.symbol)) %>%
  mutate(primary_cluster_id = sub("^plasmid_", "", contig_bin)) %>%
  group_by(sample, primary_cluster_id) %>%
  summarise(
    n_amr_hits  = n(),
    amr_genes   = paste(sort(unique(Element.symbol)), collapse = "; "),
    amr_classes = paste(sort(unique(Class[!is.na(Class)])), collapse = "; "),
    .groups     = "drop"
  )

# Join on both sample and primary_cluster_id — only bins with AMR hits
plasmid_bin_amr <- bin_size %>%
  inner_join(plasmid_amr, by = c("sample", "primary_cluster_id")) %>%
  arrange(sample, desc(total_bp))

kable(
  plasmid_bin_amr[, c("sample", "primary_cluster_id", "rep_type", "mobility",
                       "n_contigs", "total_kb", "n_amr_hits", "amr_genes", "amr_classes")],
  caption   = "Plasmid bins with AMR gene detections — sorted by size within sample",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "800px")
```




---

## Table 2 — Chromosome Contigs with AMR Genes

One row per sample, one column per unique AMR gene (hit count).
Chromosome-designated contigs may represent misclassified plasmids in short-read assemblies.
```{r chromosome-amr-table}

chrom_amr <- amr[amr$molecule_type == "chromosome", ] %>%
  filter(!is.na(Element.symbol))

# Wide format: one row per sample, one column per unique AMR gene (counts)
chrom_amr_wide <- chrom_amr %>%
  group_by(sample, Element.symbol) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from  = Element.symbol,
    values_from = count,
    values_fill = 0
  ) %>%
  arrange(sample)

kable(
  chrom_amr_wide,
  caption   = "AMR gene counts on chromosome-designated contigs per sample",
  row.names = FALSE
) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")

```

## Summary of plasmid AMR hits on both plasmid designated contigs OR chromosme designated contigs

```{r}

cat("Samples with plasmid AMR (linked on the same contigs):    ", length(unique(plasmid_bin_amr$sample)), "\n")
cat("Samples with Chromosomal AMR (Examine samples for genes that could be on plasmids):      ", length(unique(chrom_amr$sample)), "\n")
```
---

## Literature-Based Flagging

 **Placeholder** — add gene symbols or classes of interest here for manual review 
 prioritization. Examples: carbapenemases (blaNDM, blaKPC, blaOXA-48), ESBLs (blaCTX-M), 
 colistin (mcr).
```{r}


# Add gene symbols to flag here
flag_genes <- c(
  # "blaNDM", "blaKPC", "blaOXA-48", "blaCTX-M", "mcr"
)

if (length(flag_genes) > 0) {
  plasmid_bin_amr$priority_flag <- sapply(plasmid_bin_amr$amr_genes, function(g) {
    any(sapply(flag_genes, function(f) grepl(f, g, ignore.case = TRUE)))
  })
  chrom_amr_wide$priority_flag <- sapply(chrom_amr_wide$sample, function(s) {
    hits <- chrom_amr$Element.symbol[chrom_amr$sample == s]
    any(sapply(flag_genes, function(f) any(grepl(f, hits, ignore.case = TRUE))))
  })
} else {
  message("No literature flags defined — skipping flagging step.")
}

```

---

## Auspice Metadata Output

Wide-format dataframe for Auspice visualization. One row per sample, one column per
unique plasmid or chromosome AMR fingerprint. Column names are constructed as:
`{molecule_type}_{primary_cluster_id}_{rep_type}_{amr_genes_sorted}`.
Cell values are total coverage (in bp) of that bin in that sample.
AMR genes are sorted alphabetically within each fingerprint so the same plasmid
produces the same column name across different samples.
```{r}

# --- Plasmid columns ---
# Fingerprint: plasmid_{primary_cluster_id}_{rep_type}_{sorted_amr_genes}
plasmid_auspice <- plasmid_bin_amr %>%
  mutate(
    rep_type_clean = ifelse(rep_type == "" | is.na(rep_type), "-", rep_type),
    fingerprint    = paste("plasmid", primary_cluster_id, rep_type_clean, amr_genes, sep = "_")
  ) %>%
  select(sample, fingerprint, total_bp) %>%
  pivot_wider(
    names_from  = fingerprint,
    values_from = total_bp,
    values_fill = 0
  )

# --- Chromosome columns ---
# Fingerprint: chromosome_chromosome_-_{gene}
chrom_auspice <- chrom_amr %>%
  group_by(sample, Element.symbol) %>%
  summarise(
    # Use total bp of contigs carrying this gene as the value
    hit_count = n(),
    .groups   = "drop"
  ) %>%
  mutate(fingerprint = paste("chromosome", "chromosome", "-", Element.symbol, sep = "_")) %>%
  select(sample, fingerprint, hit_count) %>%
  pivot_wider(
    names_from  = fingerprint,
    values_from = hit_count,
    values_fill = 0
  )

# --- Combine into one wide dataframe ---
all_samples <- data.frame(sample_id = unique(contigs$sample))

auspice_df <- all_samples %>%
  left_join(plasmid_auspice, by = c("sample_id" = "sample")) %>%
  left_join(chrom_auspice,   by = c("sample_id" = "sample"))

# Replace NAs with 0 for samples with no hits
auspice_df[is.na(auspice_df)] <- 0

# Write out
write.csv(auspice_df, "plasmid_triage_auspice_metadata.csv", row.names = FALSE)

cat("Auspice metadata written to plasmid_triage_auspice_metadata.csv\n")
cat("Rows (samples):         ", nrow(auspice_df), "\n")
cat("Total columns:          ", ncol(auspice_df), "\n")
cat("Plasmid columns:        ", ncol(plasmid_auspice) - 1, "\n")
cat("Chromosome AMR columns: ", ncol(chrom_auspice) - 1, "\n")
```
