---
title: "Plasmid Triage Analysis"
subtitle: "MOB-suite + AMRFinderPlus Combined Output"
author: "Shawn Hawken"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This notebook combines MOB-recon contig classification with AMRFinderPlus resistance
gene calls, linking them by contig ID to answer questions like:

- Which Inc groups are carrying which resistance genes?
- Are carbapenemase genes on conjugative plasmids?
- Which plasmid bins carry multiple resistance genes?

---

## Load Libraries

```{r libraries}
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

## Load Data

> **Edit the file paths below** to point to your combined output files.

```{r load-data}
contigs <- read.delim(
  "../combined_contig_report.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

amr <- read.delim(
  "../combined_amrfinder.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)
```

**Note:** R replaces spaces and special characters in column names with `.` when
reading tab-delimited files — e.g. `Contig id` becomes `Contig.id` and
`Gene symbol` becomes `Gene.symbol`.

---

## Filter to Plasmid Contigs

```{r filter-plasmids}
plasmids <- contigs[contigs$molecule_type == "plasmid", ]

cat("Total contigs:", nrow(contigs), "\n")
cat("Plasmid contigs:", nrow(plasmids), "\n")
cat("Samples:", length(unique(plasmids$sample)), "\n")
```

---

## Join AMR Hits to Plasmid Metadata

```{r join-data}
plasmid_amr <- merge(
  plasmids, amr,
  by.x  = c("sample", "contig_id"),
  by.y  = c("sample", "Contig.id"),
  all.x = TRUE
)
```

---

## Summary Tables and figures

```{r}
# Which Inc groups carry which resistance classes?
table(plasmid_amr$rep_type, plasmid_amr$Class)

# How many plasmid bins per sample?
aggregate(primary_cluster_id ~ sample, data=plasmids, FUN=function(x) length(unique(x)))

# Total plasmid size per bin (sum contigs within each bin)
aggregate(contig_size ~ sample + primary_cluster_id, data=plasmids, FUN=sum)
```


