---
title: "Plasmid Triage Analysis"
subtitle: "MOB-suite + AMRFinderPlus Combined Output"
author: "Shawn Hawken"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Overview

This notebook combines MOB-recon contig classification with AMRFinderPlus resistance
gene calls, linking them by contig ID to answer questions like:

- Which Inc groups are carrying which resistance genes?
- Are carbapenemase genes on conjugative plasmids?
- Which plasmid bins carry multiple resistance genes?

---

## Load Libraries

```{r libraries}
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

## Load Data

> **Edit the file paths below** to point to your combined output files.

```{r load-data}
contigs <- read.delim(
  "../combined_contig_report.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

amr <- read.delim(
  "../combined_amrfinder.tsv",
  sep              = "\t",
  header           = TRUE,
  stringsAsFactors = FALSE
)

SIZE_COL <- "size"

# Filter to plasmid contigs
plasmids <- contigs[contigs$molecule_type == "plasmid", ]

# Strip plasmid_ prefix from AMRFinder bin column to match primary_cluster_id
amr$primary_cluster_id <- sub("^plasmid_", "", amr$plasmid_bin)



```

---

# summarize data

```{r filter-plasmids}


cat("Total contigs:", nrow(contigs), "\n")
cat("Plasmid contigs:", nrow(plasmids), "\n")
cat("Samples:", length(unique(plasmids$sample)), "\n")
```

---

## Join AMR Hits to Plasmid Metadata

```{r join-data}
# Summarize MOB-recon to bin level
bin_size <- aggregate(plasmids[[SIZE_COL]] ~ sample + primary_cluster_id, data=plasmids, FUN=sum)
names(bin_size)[3] <- "total_bp"
bin_size$total_kb  <- round(bin_size$total_bp / 1000, 2)

bin_size$n_contigs <- as.integer(tapply(
  plasmids[[SIZE_COL]],
  paste(plasmids$sample, plasmids$primary_cluster_id),
  length
)[paste(bin_size$sample, bin_size$primary_cluster_id)])

bin_size$rep_type <- tapply(
  plasmids$rep_type.s.,
  paste(plasmids$sample, plasmids$primary_cluster_id),
  function(x) paste(unique(x[!is.na(x) & x != "-"]), collapse="; ")
)[paste(bin_size$sample, bin_size$primary_cluster_id)]

bin_size$mobility <- tapply(
  plasmids$predicted_mobility,
  paste(plasmids$sample, plasmids$primary_cluster_id),
  function(x) paste(unique(x[!is.na(x) & x != "-"]), collapse="; ")
)[paste(bin_size$sample, bin_size$primary_cluster_id)]

# Summarize AMRFinder to bin level
amr_hits <- amr[!is.na(amr$Gene.symbol), ]

bin_amr <- aggregate(
  cbind(genes=Gene.symbol, classes=Class, subclasses=Subclass) ~ sample + primary_cluster_id,
  data = amr_hits,
  FUN  = function(x) paste(unique(x[!is.na(x)]), collapse="; ")
)
bin_amr$n_amr_hits <- as.integer(tapply(
  amr_hits$Gene.symbol,
  paste(amr_hits$sample, amr_hits$primary_cluster_id),
  length
)[paste(bin_amr$sample, bin_amr$primary_cluster_id)])

# Join
bin_combined <- merge(bin_size, bin_amr, by=c("sample", "primary_cluster_id"), all.x=TRUE)
bin_combined$has_amr <- !is.na(bin_combined$n_amr_hits)
bin_combined <- bin_combined[order(bin_combined$sample, -bin_combined$total_bp), ]
```


### Summarize plasmid bins with AMR genes

Full combined table — one row per plasmid bin with size and AMR cargo.
Bins with no AMR hits are included with empty gene columns.

```{r cargo-table}
display_cols <- c("sample", "primary_cluster_id", "rep_type", "mobility",
                  "n_contigs", "total_kb", "n_amr_hits", "genes", "classes", "subclasses")

kable(bin_combined[, display_cols],
      caption   = "Plasmid bin size and AMR gene cargo",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "400px")


```


### subset to only plasmid bins with AMR data

```{r}
AMR_bin<-bin_combined[bin_combined$has_amr==TRUE,]

# make reptype_gene variable to be able to find all unique combos of replication types carrying the same genes (ie unique plasmids)
AMR_bin$rep_type_genes<-paste(AMR_bin$rep_type, AMR_bin$genes,sep = '_')

# make reptype_gene variable to be able to find all unique combos of plasmid bins  carrying the same genes 
AMR_bin$bin_type_genes<-paste(AMR_bin$primary_cluster_id, AMR_bin$genes,sep = '_')


```

